# 吸烟导致癌症？ {#cancer}

```{r cancer_setup,message=FALSE,cache=FALSE}
library(tidyverse)
library(rstan)
```


从肺癌患者和无癌个体中随机抽取两组样本[^cancer]，目的是考察两组人吸烟习惯的差异。
数据如下

```{r cancer01}
cancer <- tribble(
  ~group, ~n, ~smokers,
  "Cancer patients", 86, 82,
  "Control group", 86, 72
)
cancer
```



## Two Sample Binomial Model

In implementing this model, we have just two data points (cancer patients and
control group) and a binomial sampling model, in which the population
proportions of smokers in each group appear as parameters.  Quantities of
interest such as the difference in the population proportions and the log of
the odds ratio are computed in the generated quantities section. Uniform priors
on the population proportions are used in this example.

$$
\begin{aligned}[t]
r_i &\sim \mathsf{Binomial}(n_i, \pi_i)
\end{aligned}
$$
Additionally the difference,
$$
\delta = \pi_1 - \pi_2 ,
$$
and the log-odds ratio,
$$
\lambda = \log\left(\frac{\pi_1}{1 - \pi_1}\right) - \log \left( \frac{\pi_2}{1 - \pi_2} \right) ,
$$

It places uniform priors (Beta priors) are placed on $\pi$,
$$
\begin{aligned}
\pi_i &\sim \mathsf{Beta}(1, 1)
\end{aligned}
$$

The difference between and log odds ratio are defined in the `generated quantities` block.



The Stan model for this is:
```{r}
cancer_mod1 <- "
data {
  int<lower = 0> r[2];
  int<lower = 1> n[2];
  // param for beta prior on p
  vector<lower = 0>[2] p_a;
  vector<lower = 0>[2] p_b;
}
parameters {
  vector<lower = 0, upper = 1>[2] p;
}
model {
  p ~ beta(p_a, p_b);
  r ~ binomial(n, p);
}
generated quantities {
  real delta;
  int delta_up;
  real lambda;
  int lambda_up;

  delta = p[1] - p[2];
  delta_up = delta > 0;
  lambda = logit(p[1]) - logit(p[2]);
  lambda_up = lambda > 0;

}
"


cancer_data <- list(
  r = cancer$smokers,
  n = cancer$n,
  # beta prior on pi
  p_a = rep(1, 2),
  p_b = rep(1, 2)
)


cancer_fit1 <- stan(model_code = cancer_mod1, data = cancer_data)
```



```{r}
cancer_fit1
```

## Binomial Logit Model of the Difference

An alternative parameterization directly models the difference in the population proportion.

$$
\begin{aligned}[t]
r_i &\sim \mathsf{Binomial}(n_i, \pi_i) \\
\pi_1 &= \frac{1}{1 + \exp(-(\alpha + \beta)} \\
\pi_2 &= \frac{1}{1 + \exp(-\alpha))}
\end{aligned}
$$
The parameters $\alpha$ and $\beta$ are given weakly informative priors on the log-odds scale,
$$
\begin{aligned}
\alpha &\sim N(0, 10)\\
\beta &\sim N(0, 2.5)
\end{aligned}
$$

```{r cancer_mod2,results='hide',cache.extra=tools::md5sum("stan/cancer2.stan")}

stan_program <- "
data {
  int<lower = 0> r[2];
  int<lower = 1> n[2];
  // param for beta prior on p
  real a_loc;
  real<lower = 0.> a_scale;
  real b_loc;
  real<lower = 0.> b_scale;
}
parameters {
  real a;
  real b;
}
transformed parameters {
  vector<lower = 0., upper = 1.>[2] p;
  p[1] = inv_logit(a + b);
  p[2] = inv_logit(a);
}
model {
  a ~ normal(a_loc, a_scale);
  b ~ normal(a_loc, b_scale);
  r ~ binomial(n, p);
}
generated quantities {
  real delta;
  int delta_up;
  real lambda;
  int lambda_up;

  delta = p[1] - p[2];
  delta_up = delta > 0;
  lambda = logit(p[1]) - logit(p[2]);
  lambda_up = lambda > 0;

}

```


Re-use `r` and `n` values from `cancer_data`, but add the appropriate values for the prior distributions.
```{r cancer_data2}
cancer_data2 <- within(cancer_data, {
  p_a <- p_b <- NULL
  a_loc <- b_loc <- 0
  a_scale <- 10
  b_scale <- 2.5
})
```

Sample from the model:
```{r cancer_fit2,results='hide'}
cancer_fit2 <- sampling(cancer_mod2, cancer_data2)
```
```{r}
cancer_fit2
```

## Questions

1.  Expression the Binomial Logit model of the Difference as a regression
1.  What number of success and failures is a `Beta(1,1)` prior equivalent to?

[^cancer]: This example is derived from Simon Jackman,
    "[Cancer: difference in two binomial proportions](https://web-beta.archive.org/web/20070601000000*/http://jackman.stanford.edu:80/mcmc/cancer.odc)",
    *BUGS Examples,* 2007-07-24, This examples comes from @JohnsonAlbert1999a, using data from @Dorn1954a.
