# 物业经理的困惑 {#complaints}

```{r libraries, echo = FALSE}
library(tidyverse)
library(tidybayes)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```


蟑螂投诉的案例，
一方面担心投诉太多降低威望，一方面担心部署太多蟑螂诱捕器需要花费太多费用

## 数据

```{r}
pest <- readRDS("./data/pest_data.RDS")
pest
```

```{r}
pest %>% count(building_id, live_in_super)
```

## Hierarchical modeling

modeling varying intercepts for each building

```{r, warning=FALSE, message=FALSE}
stan_program <- "
functions {
  /*
  * Alternative to neg_binomial_2_log_rng() that 
  * avoids potential numerical problems during warmup
  */
  int neg_binomial_2_log_safe_rng(real eta, real phi) {
    real gamma_rate = gamma_rng(phi, phi / exp(eta));
    if (gamma_rate >= exp(20.79))
      return -9;     
    return poisson_rng(gamma_rate);
  }
}
data {
  int<lower=1> N;
  int complaints[N];
  int traps[N];
  int n_building;
  int building[N];
  vector<lower=0,upper=1>[N] live_in_super;
  vector[N] log_sq_foot;
}
parameters {
  vector[n_building] alpha;
  real beta;
  real beta_super;
  real<lower=0> inv_phi;
}
transformed parameters {
  real phi = inv(inv_phi);
}
model {
  vector[N] lambda;
  for (i in 1:N) {
    lambda[i] = alpha[building[i]] + beta * traps[i] + beta_super * live_in_super[i] + log_sq_foot[i];
  }

  
  for (i in 1:N) {
    target += neg_binomial_2_log_lpmf(complaints[i] | lambda[i], phi);
  }
  
  alpha ~ normal(log(4), 1);
  beta ~ normal(-0.25, 1);
  beta_super ~ normal(-0.5, 1);
  inv_phi ~ normal(0, 1); 

  
} 
generated quantities {
  int y_rep[N];
  for (n in 1:N) {
    y_rep[n] = neg_binomial_2_log_safe_rng(alpha[building[n]] + beta * traps[n] + beta_super * live_in_super[n] + log_sq_foot[n], phi);
  }
}
"



stan_data <- pest %>% 
  select(complaints, traps, building_id, live_in_super, total_sq_foot) %>% 
  mutate(log_sq_foot  = log(total_sq_foot/1e4)) %>% 
  mutate(building_id  = as_factor(building_id)) %>% #compose_data() convert factors into `int` for grouping
  
  tidybayes::compose_data(
  N             = nrow(.), 
  complaints    = complaints,
  traps         = traps,
  live_in_super = live_in_super,
  log_sq_foot   = log_sq_foot,
  n_building    = n_distinct(building_id),
  building      = building_id
 )


fit_pest_Hierarchical <- stan(model_code = stan_program, data = stan_data)
```




$$
\text{complaints}_{b,t} \sim \text{Neg-Binomial}(\lambda_{b,t}, \phi) \\
\lambda_{b,t}  = \exp{(\eta_{b,t})} \\
\eta_{b,t} = \mu_b + \beta \, {\rm traps} + \text{log}\_\textrm{sq}\_\textrm{foot}\\
\mu_b \sim \text{normal}(\alpha + \texttt{building}\_\textrm{data} \, \zeta, \,\sigma_{\mu})
$$











```{r, warning=FALSE, message=FALSE}
stan_program <- "
functions {
  /*
  * Alternative to neg_binomial_2_log_rng() that 
  * avoids potential numerical problems during warmup
  */
  int neg_binomial_2_log_safe_rng(real eta, real phi) {
    real gamma_rate = gamma_rng(phi, phi / exp(eta));
    if (gamma_rate >= exp(20.79))
      return -9;      
    return poisson_rng(gamma_rate);
  }
}

data {
  int<lower=1> N;
  int<lower=0> complaints[N]; 
  int traps[N];
  int<lower=1> n_building;
  int<lower=1, upper=n_building> building[N];
  vector[N] log_sq_foot;
  matrix[n_building,4] building_data;
}


parameters {
  real<lower=0> inv_phi;   // 1/phi (easier to think about prior for 1/phi instead of phi)
  real beta;               // coefficient on traps
  vector[n_building] mu;   // buildings-specific intercepts
  real<lower=0> sigma_mu;  // sd of building-specific intercepts
  real alpha;              // intercept of model for mu
  vector[4] zeta;          // coefficients on building-level predictors in model for mu 
}
transformed parameters {
  real phi = inv(inv_phi);
}


model {

  sigma_mu ~ normal(0, 1);
  alpha ~ normal(log(4), 1);
  zeta ~ normal(0, 1);      // could also use informative priors on the different elements
  beta ~ normal(-0.25, 1);
  inv_phi ~ normal(0, 1);

  
  //for (j in 1:n_building) {
  //  mu[j] ~ normal(alpha + building_data[j, ]  * zeta, sigma_mu);
  //}
  mu ~ normal(alpha + building_data * zeta, sigma_mu);
  
  for (i in 1:N) {
    target += neg_binomial_2_log_lpmf(complaints[i] | mu[building[i]] + beta * traps[i] + log_sq_foot[i], phi);
  }
  

} 
generated quantities {
  int y_rep[N];
  for (n in 1:N) {
    y_rep[n] = neg_binomial_2_log_safe_rng(mu[building[n]] + beta * traps[n] + log_sq_foot[n], phi);
  }
}
"
```


```{r prep-data}
N_buildings <- length(unique(pest$building_id))
N_months    <- length(unique(pest$date))

#
building_data <- pest %>%
  mutate(
    building_fac = factor(building_id, levels = unique(building_id)),
    building_idx = as.integer(building_fac),
    ids = rep(1:N_months, N_buildings),
    month_idx = lubridate::month(date)
  ) %>% 
    select(building_idx, 
           live_in_super, 
           age_of_building,
           total_sq_foot, 
           average_tenant_age, 
           monthly_average_rent) %>%
    distinct() %>%
    arrange(building_idx) %>%
    select(-building_idx) %>%
    mutate(
    across(everything(), scale, scale = FALSE)
    ) %>% 
    mutate( # scale by constants
      age_of_building = age_of_building / 10,
      total_sq_foot = total_sq_foot / 10000,              # not used
      average_tenant_age = average_tenant_age / 10,
      monthly_average_rent = monthly_average_rent / 1000
    ) %>%
    as.matrix()

building_data
building_data[, -3]
```




```{r}
stan_data <- pest %>% 
  select(complaints, traps, building_id, total_sq_foot) %>% 
  mutate(log_sq_foot  = log(total_sq_foot/1e4)) %>% 
  mutate(building_id  = as_factor(building_id)) %>% 
  
  
  tidybayes::compose_data(
  N             = nrow(.), 
  complaints    = complaints,
  traps         = traps,
  log_sq_foot   = log_sq_foot,
  n_building    = n_distinct(building_id),
  building      = building_id,
  building_data = building_data[,-3],
 )

fitted_model_NB_hier <- stan(model_code = stan_program, data = stan_data)
```


有效样本都很低， 

```{r}
fitted_model_NB_hier %>% 
  print(pars = c('sigma_mu','beta','alpha','phi','mu'))
```

我们先看trace plots

```{r}
# use as.array to keep the markov chains separate for trace plots
bayesplot::mcmc_trace(as.array(fitted_model_NB_hier, pars = 'sigma_mu'),
  np = nuts_params(fitted_model_NB_hier), window = c(500,1000)
)
```
Looks as if the divergent parameters, the little red bars underneath
the traceplots correspond to samples where the sampler gets stuck at
one parameter value for $\sigma_\mu$.



### want this to look more like a funnel than a cloud

What we have here is a cloud-like shape, with most of the divergences
clustering towards the bottom. We'll see a bit later that we actually
want this to look more like a funnel than a cloud.

```{r}
# assign to object so we can compare to another plot later
scatter_with_divs <- bayesplot::mcmc_scatter(as.array(fitted_model_NB_hier),
  pars = c("mu[4]", 'sigma_mu'),
  transform = list('sigma_mu' = "log"), np = nuts_params(fitted_model_NB_hier)
)
scatter_with_divs
```

这是我们期待的漏斗图（因为是我们模拟的）

```{r}
d_sim <- tibble(
    log_sigma = rnorm(1000, mean = 0, sd = 1)
) %>%
  mutate(
    theta = map_dbl(log_sigma, ~ rnorm(1, mean = 0, sd = exp(.x)))
  )
d_sim

d_sim %>% 
  ggplot(aes(x = log_sigma, y = theta)) +
  geom_point()
```

```{r}
N_sims <- 1000
log_sigma <- rep(NA, N_sims)
theta <- rep(NA, N_sims)
for (j in 1:N_sims) {
  log_sigma[j] <- rnorm(1, mean = 0, sd = 1)
  theta[j] <- rnorm(1, mean = 0, sd = exp(log_sigma[j]))
}
draws <- cbind("mu" = theta, "log(sigma_mu)" = log_sigma)
bayesplot::mcmc_scatter(draws)
```


```{r}
parcoord_with_divs <-
  bayesplot::mcmc_parcoord(as.array(fitted_model_NB_hier, pars = c("sigma_mu", "mu")),
  np = nuts_params(fitted_model_NB_hier))
parcoord_with_divs
```



## hier_NB_regression_ncp
上面的分析，发现有效样本量比较低，原因是sigma_mu 太小，所以作者这里Reparameterizing,

$$
\text{complaints}_{b,t} \sim \text{Neg-Binomial}(\lambda_{b,t}, \phi) \\
\lambda_{b,t}  = \exp{(\eta_{b,t})} \\
\eta_{b,t} = \mu_b + \beta \, {\rm traps} + \text{log}\_\textrm{sq}\_\textrm{foot}\\
\mu_b \sim \text{normal}(\alpha + \texttt{building}\_\textrm{data} \, \zeta, \,\sigma_{\mu})
$$











```{r, warning=FALSE, message=FALSE}
stan_program <- "
functions {
  /*
  * Alternative to neg_binomial_2_log_rng() that 
  * avoids potential numerical problems during warmup
  */
  int neg_binomial_2_log_safe_rng(real eta, real phi) {
    real gamma_rate = gamma_rng(phi, phi / exp(eta));
    if (gamma_rate >= exp(20.79))
      return -9;      
    return poisson_rng(gamma_rate);
  }
}

data {
  int<lower=1> N;
  int<lower=0> complaints[N]; 
  int traps[N];
  int<lower=1> n_building;
  int<lower=1, upper=n_building> building[N];
  vector[N] log_sq_foot;
  matrix[n_building,4] building_data;
}


parameters {
  real<lower=0> inv_phi;   // 1/phi (easier to think about prior for 1/phi instead of phi)
  real beta;               // coefficient on traps
  real<lower=0> sigma_mu;  // sd of building-specific intercepts
  real alpha;              // intercept of model for mu
  vector[4] zeta;          // coefficients on building-level predictors in model for mu 
  vector[n_building] mu_raw;
}
transformed parameters {
  real phi = inv(inv_phi);
  vector[n_building] mu;
  mu = alpha + building_data * zeta + sigma_mu * mu_raw;
}


model {

  sigma_mu ~ normal(0, 1);
  alpha ~ normal(log(4), 1);
  zeta ~ normal(0, 1);      // could also use informative priors on the different elements
  beta ~ normal(-0.25, 1);
  inv_phi ~ normal(0, 1);
  mu_raw ~ normal(0, 1);    // implies mu ~ normal(alpha + building_data * zeta, sigma_mu)
  
  
  for (i in 1:N) {
    target += neg_binomial_2_log_lpmf(complaints[i] | mu[building[i]] + beta * traps[i] + log_sq_foot[i], phi);
  }
  

} 
generated quantities {
  int y_rep[N];
  for (n in 1:N) {
    y_rep[n] = neg_binomial_2_log_safe_rng(mu[building[n]] + beta * traps[n] + log_sq_foot[n], phi);
  }
}
"
```





```{r}
N_buildings <- length(unique(pest$building_id))
N_months    <- length(unique(pest$date))

#
building_data <- pest %>%
  mutate(
    building_fac = factor(building_id, levels = unique(building_id)),
    building_idx = as.integer(building_fac),
    ids = rep(1:N_months, N_buildings),
    month_idx = lubridate::month(date)
  ) %>% 
    select(building_idx, 
           live_in_super, 
           age_of_building,
           total_sq_foot, 
           average_tenant_age, 
           monthly_average_rent) %>%
    distinct() %>%
    arrange(building_idx) %>%
    select(-building_idx) %>%
    mutate(
    across(everything(), scale, scale = FALSE)
    ) %>% 
    mutate( # scale by constants
      age_of_building = age_of_building / 10,
      total_sq_foot = total_sq_foot / 10000,              # not used
      average_tenant_age = average_tenant_age / 10,
      monthly_average_rent = monthly_average_rent / 1000
    ) %>%
    as.matrix()


stan_data <- pest %>% 
  select(complaints, traps, building_id, total_sq_foot) %>% 
  mutate(log_sq_foot  = log(total_sq_foot/1e4)) %>% 
  mutate(building_id  = as_factor(building_id)) %>% 
  
  
  tidybayes::compose_data(
  N             = nrow(.), 
  complaints    = complaints,
  traps         = traps,
  log_sq_foot   = log_sq_foot,
  n_building    = n_distinct(building_id),
  building      = building_id,
  building_data = building_data[,-3],
 )

fitted_model_NB_hier_ncp <- stan(model_code = stan_program, data = stan_data)
```


再看看有效样本

```{r}
fitted_model_NB_hier_ncp %>% 
  print(pars = c('sigma_mu','beta','alpha','phi','mu'))
```


haha 有效样本改进很大耶
```{r}
scatter_no_divs <- bayesplot::mcmc_scatter(as.array(fitted_model_NB_hier_ncp),
  pars = c("mu[4]", 'sigma_mu'), transform = list('sigma_mu' = "log"),
  np = nuts_params(fitted_model_NB_hier_ncp))
bayesplot::bayesplot_grid(scatter_with_divs, scatter_no_divs,
               grid_args = list(ncol = 2), ylim = c(-11, 1))
```

```{r}
parcoord_no_divs <- bayesplot::mcmc_parcoord(
  as.array(fitted_model_NB_hier_ncp, pars = c("sigma_mu", "mu")),
  np = nuts_params(fitted_model_NB_hier_ncp)
)
bayesplot::bayesplot_grid(parcoord_with_divs, parcoord_no_divs, ylim = c(-3, 3))
```


```{r sims-full-hier}
sims_NB_hier_ncp <-
  rstan::extract(fitted_model_NB_hier_ncp, pars = c('y_rep','inv_phi'))
```

The marginal plot, again:

```{r ppc-full-hier}
y_rep <- as.matrix(fitted_model_NB_hier_ncp, pars = "y_rep")
bayesplot::ppc_dens_overlay(stan_data$complaints, y_rep[1:200,])
```

Predictions by number of bait stations:

```{r}
y_rep <- as.matrix(fitted_model_NB_hier_ncp, pars = "y_rep")
bayesplot::ppc_intervals(y = stan_data$complaints, yrep = y_rep,
              x = stan_data$traps) +
  labs(x = "Number of bait stations", y = "Number of complaints")
```




# Varying intercepts and varying slopes  {-}


```{r}
stan_dat_hier <- readRDS('data/pest_data_longer_stan_dat.RDS')
stan_dat_hier %>% str()
```

这个数据比之前的要丰富些，因此，我们需要重新组建一下，弄成

```{r}
stan_data <- list(
     N = 360,
     complaints    = stan_dat_hier$complaints,
     traps         = stan_dat_hier$traps,
     log_sq_foot   = stan_dat_hier$log_sq_foot,
     building_id   = stan_dat_hier$building_idx,
     n_building    = unique(stan_dat_hier$building_idx),
     building_data = stan_dat_hier$building_data
)
```


模型需要扩展下

$$
\text{complaints}_{b,t} \sim \text{Neg-Binomial}(\lambda_{b,t}, \phi)
\\
\lambda_{b,t} = \exp{(\eta_{b,t})}
\\
\eta_{b,t} = \mu_b + \kappa_b \, \texttt{traps}_{b,t}
             + \text{log}\_\textrm{sq}\_\textrm{foot}_b
\\
\mu_b \sim \text{normal}(\alpha + \texttt{building}\_\textrm{data} \, \zeta,
                         \sigma_{\mu}) \\
\kappa_b \sim \text{normal}(\beta + \texttt{building}\_\textrm{data} \, \gamma,
                            \sigma_{\kappa})
$$



```{r, warning=FALSE, message=FALSE}
stan_program <- "
functions {
  /*
  * Alternative to neg_binomial_2_log_rng() that 
  * avoids potential numerical problems during warmup
  */
  int neg_binomial_2_log_safe_rng(real eta, real phi) {
    real gamma_rate = gamma_rng(phi, phi / exp(eta));
    if (gamma_rate >= exp(20.79))
      return -9;      
    return poisson_rng(gamma_rate);
  }
}

data {
  int<lower=1> N;
  int<lower=0> complaints[N]; 
  int traps[N];
  int<lower=1> n_building;
  int<lower=1, upper=n_building> building[N];
  vector[N] log_sq_foot;
  matrix[n_building,4] building_data;
}


parameters {
 
  real alpha;
  real beta;               
  real<lower=0> sigma_mu;  
  real<lower=0> sigma_kappa;  
  vector[n_building] mu_raw;
  vector[n_building] kappa_raw;
  
  vector[4] zeta;          
  vector[4] gamma;   
  real<lower=0> inv_phi;  
}
transformed parameters {
  real phi = inv(inv_phi);
  vector[n_building] mu = alpha + building_data * zeta + sigma_mu * mu_raw;
  vector[n_building] kappa = beta + building_data * gamma + sigma_kappa * kappa_raw;
}


model {

  alpha ~ normal(log(4), 1);
  beta ~ normal(-0.25, 1);
  sigma_mu ~ normal(0, 1);
  sigma_kappa ~ normal(0, 1);
  mu_raw ~ normal(0,1);
  kappa_raw ~ normal(0, 1);
  zeta ~ normal(0, 1);      
  gamma ~ normal(0, 1);      
  inv_phi ~ normal(0, 1);

  
  for (i in 1:N) {
    target += neg_binomial_2_log_lpmf(complaints[i] | mu[building[i]] + kappa[building[i]] * traps[i] + log_sq_foot[i], phi);
  }
  

} 
generated quantities {
  int y_rep[N];
  for (n in 1:N) {
    y_rep[n] = neg_binomial_2_log_safe_rng(mu[building[n]] + kappa[building[n]] * traps[n] + log_sq_foot[n], phi);
  }
}
"
```


```{r}
stan_data <- list(
     N = 360,
     complaints    = stan_dat_hier$complaints,
     traps         = stan_dat_hier$traps,
     log_sq_foot   = stan_dat_hier$log_sq_foot,
     building      = stan_dat_hier$building_idx,
     n_building    = length(unique(stan_dat_hier$building_idx)),
     building_data = stan_dat_hier$building_data
)

fitted_model_NB_hier_slopes <- stan(model_code = stan_program, data = stan_data)
```

```{r}
bayesplot::mcmc_hist(as.matrix(fitted_model_NB_hier_slopes, pars = "sigma_kappa"),
          binwidth = 0.005
)
```

```{r}
print(fitted_model_NB_hier_slopes,
      pars = c('kappa','beta','alpha','phi','sigma_mu','sigma_kappa','mu'))
```

```{r}
bayesplot::mcmc_hist(as.matrix(fitted_model_NB_hier_slopes, pars = "beta"),
          binwidth = 0.005
)
```


```{r ppc-full-hier-slopes}
y_rep <- as.matrix(fitted_model_NB_hier_slopes, pars = "y_rep")
bayesplot::ppc_dens_overlay(y = stan_dat_hier$complaints, yrep = y_rep[1:200,]
)
```