# 身高体重 {#height-weight}

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidybayes)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```


## 线性模型

从最简单的线性模式开始
$$
y_n = \alpha + \beta x_n + \epsilon_n \quad \text{where}\quad
\epsilon_n \sim \operatorname{normal}(0,\sigma).
$$

等价于

$$
y_n - (\alpha + \beta X_n) \sim \operatorname{normal}(0,\sigma),
$$

进一步等价

$$
y_n \sim \operatorname{normal}(\alpha + \beta X_n, \, \sigma).
$$


```{r bayes-simuate, eval=FALSE}
alpha_real <- 10
beta_real  <- 3
sigma_real <- 2

df <- tibble(
  x = runif(30, 1, 8),
  y = rnorm(30, alpha_real + beta_real * x, sd = sigma_real)
)
```



```{r, eval=FALSE}
stan_program <- "
data {
  int<lower=0> N;
  vector[N] x;
  vector[N] y;
}
parameters {
  real alpha;
  real beta;
  real<lower=0> sigma;
}
model {
  y ~ normal(alpha + beta * x, sigma);
}
generated quantities {
  vector[N] y_rep; 
  for (n in 1:N) 
    y_rep[n] = normal_rng(alpha + beta * x[n], sigma);
}

"

stan_data <- df %>%
  tidybayes::compose_data(
    N = nrow(.),
    x = x,
    y = y
  )

fit_normal <- stan(model_code = stan_program, data = stan_data)
```

```{r include=FALSE}
# 运行stan代码，导致渲染bookdown报错，不知道为什么，先用这边笨办法凑合吧
#
#save(fit_normal, 
#     stan_data, 
#     alpha_real, 
#     beta_real, 
#     sigma_real, 
#     file = here::here("stan", "stan_data_normal.Rdata")
#     ) 

load(here::here("stan", "stan_data_normal.Rdata"))
```

### 模型输出

```{r}
fit_normal
```



### 模型评估
```{r}
rstan::traceplot(fit_normal, pars = c("alpha", "beta", "sigma"))
```


```{r, eval=FALSE}
rstan::extract(fit_normal, par = c("alpha", "beta"))
rstan::extract(fit_normal, par = "alpha")$alpha
rstan::extract(fit_normal, par = "beta")$beta
```





```{r, eval=FALSE}
fit_normal %>% 
  tidybayes::gather_draws(alpha, beta) %>% 
  ggplot(aes(x = .value, y = as_factor(.variable)) ) +
  ggdist::stat_halfeye() +
  geom_vline(xintercept = c(alpha_real, beta_real)) 
```



事实上，`bayesplot`宏包提供了大量模型评估函数，大爱!!

```{r, message=FALSE, results=FALSE}
true_alpha_beta <- c(alpha_real, beta_real, sigma_real)

posterior_alpha_beta <- 
  as.matrix(fit_normal, pars = c('alpha','beta', 'sigma'))
bayesplot::mcmc_recover_hist(posterior_alpha_beta, true = true_alpha_beta)
```


```{r}
y_rep <- as.matrix(fit_normal, pars = "y_rep")
bayesplot::ppc_dens_overlay(y = stan_data$y, yrep = y_rep[1:200, ])
```



```{r bayes-09}
y_rep <- as.matrix(fit_normal, pars = "y_rep")
bayesplot::ppc_intervals(y = stan_data$y, yrep = y_rep, x = stan_data$x) 
```



## bayesian workflow




## 参考资料

- https://mc-stan.org/
- https://github.com/jgabry/bayes-workflow-book
- https://github.com/XiangyunHuang/masr/
- https://github.com/ASKurz/Statistical_Rethinking_with_brms_ggplot2_and_the_tidyverse_2_ed/
- 《Regression and Other Stories》, Andrew Gelman, Cambridge University Press. 2020
- 《A Student's Guide to Bayesian Statistics》,  Ben Lambert, 2018
- 《Statistical Rethinking: A Bayesian Course with Examples in R and STAN》 ( 2nd Edition), by Richard McElreath, 2020
- 《Bayesian Data Analysis》, Third Edition, 2013
- 《Doing Bayesian Data Analysis: A Tutorial with R, JAGS, and Stan》  (2nd Edition) John Kruschke, 2014
- 《Bayesian Models for Astrophysical Data: Using R, JAGS, Python, and Stan》， Joseph M. Hilbe, Cambridge University Press, 2017


```{r bayes-20, echo = F}
# remove the objects
# ls() %>% stringr::str_flatten(collapse = ", ")

rm(fit_normal, y_rep, stan_data, alpha_real, beta_real, sigma_real,posterior_alpha_beta, true_alpha_beta)
```



```{r bayes-21, echo = F, message = F, warning = F, results = "hide"}
pacman::p_unload(pacman::p_loaded(), character.only = TRUE)
```

