# 咖啡等待时间 {#wait-coffe}

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(brms)
library(tidybayes)
library(bayesplot)
library(rstan)
library(patchwork)
options(mc.cores = 4)
```

# Section 14.1: Varying slopes by construction

```{r}
a       <-  3.5  # average morning wait time
b       <- -1    # average difference afternoon wait time
sigma_a <-  1    # std dev in intercepts
sigma_b <-  0.5  # std dev in slopes
rho     <- -.7   # correlation between intercepts and slopes
# the next three lines of code simply combine the terms, above
mu     <- c(a, b)
cov_ab <- sigma_a * sigma_b * rho
sigma  <- matrix(c(sigma_a^2, cov_ab, 
                   cov_ab, sigma_b^2), ncol = 2)
```



```{r, message = F, warning = F}
sigmas <- c(sigma_a, sigma_b)          # standard deviations
rho    <- matrix(c(1, rho,             # correlation matrix
                   rho, 1), nrow = 2)
# now matrix multiply to get covariance matrix
sigma <- diag(sigmas) %*% rho %*% diag(sigmas)
# how many cafes would you like?
n_cafes <- 20
set.seed(5)  # used to replicate example
vary_effects <- 
  MASS::mvrnorm(n_cafes, mu, sigma) %>% 
  data.frame() %>% 
  set_names("a_cafe", "b_cafe")
head(vary_effects)
```



```{r}
n_visits <- 10
sigma    <-  0.5  # std dev within cafes
set.seed(22)  # used to replicate example
d <-
  vary_effects %>% 
  mutate(cafe = 1:n_cafes) %>% 
  expand(nesting(cafe, a_cafe, b_cafe), visit = 1:n_visits) %>% 
  mutate(afternoon = rep(0:1, times = n() / 2)) %>% 
  mutate(mu = a_cafe + b_cafe * afternoon) %>% 
  mutate(wait = rnorm(n = n(), mean = mu, sd = sigma))
```

We might peek at the data.

```{r}
d %>%
  glimpse()
```
```{r}
d1 <- d %>% select(wait, afternoon, cafe)
d1
```


```{r}
d1 %>% write_rds("./data/cafe.rds")
```





```{r}
d <- readr::read_rds("./data/cafe.rds")
```





### The varying slopes model.

The statistical formula for our varying intercepts and slopes café model follows the form


$$
\begin{align*}
\text{wait}_i & \sim \operatorname{Normal}(\mu_i, \sigma) \\
\mu_i         & = \alpha_{\text{café}[i]} + \beta_{\text{café}[i]} \text{afternoon}_i \\
\begin{bmatrix} \alpha_\text{café} \\ \beta_\text{café} \end{bmatrix} & \sim \operatorname{MVNormal} \begin{pmatrix} \begin{bmatrix} \alpha \\ \beta \end{bmatrix}, \mathbf{S} \end{pmatrix} \\
\mathbf S     & = \begin{bmatrix} \sigma_\alpha & 0 \\ 0 & \sigma_\beta \end{bmatrix} \mathbf R \begin{bmatrix} \sigma_\alpha & 0 \\ 0 & \sigma_\beta \end{bmatrix} \\
\alpha        & \sim \operatorname{Normal}(5, 2) \\
\beta         & \sim \operatorname{Normal}(-1, 0.5) \\
\sigma        & \sim \operatorname{Exponential}(1) \\
\sigma_\alpha & \sim \operatorname{Exponential}(1) \\
\sigma_\beta  & \sim \operatorname{Exponential}(1) \\
\mathbf R     & \sim \operatorname{LKJcorr}(2),
\end{align*}
$$


感觉这里没说清楚，比如$\alpha_{cafe[i]}$ 应该是两个，一个是固定效应，一个变化效应
各自的分布是什么呢？




stan2 很简洁
```{r, warning=FALSE, message=FALSE}
stan_program <- "
data {
  int n;
  int n_cafe;
  int cafe[n];
  vector[n] afternoon;
  vector[n] wait;
}
parameters {
  vector[n_cafe] a_cafe;
  vector[n_cafe] b_cafe;
  real a;
  real b;
  vector<lower=0>[2] sigma_cafe;
  real<lower=0> sigma;
  corr_matrix[2] Rho;
}
model {
  vector[n] mu;
  vector[2] YY[n_cafe];
  vector[2] MU;
  Rho ~ lkj_corr(2);
  sigma ~ exponential(1);
  sigma_cafe ~ exponential(1);
  a ~ normal(5, 2);
  b ~ normal(-1, .5);
  MU = [a, b]';
  for (j in 1:n_cafe) {
    YY[j] = [a_cafe[j], b_cafe[j]]';
  }
  YY ~ multi_normal(MU, quad_form_diag(Rho, sigma_cafe));
  mu = a_cafe[cafe] + b_cafe[cafe] .* afternoon;
  wait ~ normal(mu, sigma);
}
"


stan_data <- compose_data(d,
 		       n_cafe = n_distinct(cafe)
 		       )


m14.1a <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m14.1a
```






用stan1的代码， 结构比较清晰
```{r, warning=FALSE, message=FALSE}
stan_program <- "
data {
  int n;               //多少观察值
  int n_cafe;          //分了多少组
  int cafe[n];
  real wait[n];
  int afternoon[n];
}
parameters {
  real a;
  real b;
  vector[n_cafe] a_cafe;
  vector[n_cafe] b_cafe;
  vector<lower=0>[2] sigma_cafe;
  real<lower=0> sigma;
  corr_matrix[2] Rho;
}
transformed parameters {
  vector[2] MU;
  vector[2] v_a_cafe_b_cafe[n_cafe];
  cov_matrix[2] SRS_sigma_cafeRho;
  MU = [a, b]';
  for (j in 1:n_cafe) v_a_cafe_b_cafe[j] = [a_cafe[j], b_cafe[j]]';
  SRS_sigma_cafeRho = quad_form_diag(Rho, sigma_cafe);
}
model {
  vector[n] mu;
  
 // priors
 target += normal_lpdf(a| 5, 2);
 target += normal_lpdf(b| -1, 0.5);
 target += exponential_lpdf(sigma| 1);
 target += exponential_lpdf(sigma_cafe| 1);
 target += lkj_corr_lpdf(Rho| 2);
 
 // linear model
 for(i in 1:n) {
   mu[i] = a_cafe[cafe[i]] + b_cafe[cafe[i]] * afternoon[i];
 }
 target += normal_lpdf(wait |mu, sigma);
 target += multi_normal_lpdf(v_a_cafe_b_cafe | MU, SRS_sigma_cafeRho);
}
"

stan_data <- d %>% 
  tidybayes::compose_data(
 		n_cafe = n_distinct(cafe)
 	)


m14.1b <- stan(model_code = stan_program, data = stan_data)

```



```{r}
m14.1b
```







我的目的是：语法简介，结构也清晰
```{r, warning=FALSE, message=FALSE}
stan_program <- "
data {
  int n;
  int n_cafe;
  int cafe[n];
  vector[n] afternoon;
  vector[n] wait;
}
parameters {
  vector[n_cafe] a_cafe;
  vector[n_cafe] b_cafe;
  real a;
  real b;
  vector<lower=0>[2] sigma_cafe;
  real<lower=0> sigma;
  corr_matrix[2] Rho;
}
transformed parameters {
  vector[2] YY[n_cafe];
  vector[2] MU;
  MU = [a, b]';
  for (j in 1:n_cafe) {
    YY[j] = [a_cafe[j], b_cafe[j]]';
  }
}
model {
  vector[n] mu;

  sigma ~ exponential(1);
  sigma_cafe ~ exponential(1);
  a ~ normal(5, 2);
  b ~ normal(-1, .5);
  Rho ~ lkj_corr(2);
  
  mu = a_cafe[cafe] + b_cafe[cafe] .* afternoon;
  
  //YY ~ multi_normal(MU, quad_form_diag(Rho, sigma_cafe));
  target += multi_normal_lpdf(YY | MU, quad_form_diag(Rho, sigma_cafe));
  wait ~ normal(mu, sigma);
}
"


stan_data <- d %>% 
  tidybayes::compose_data(
 		n_cafe = n_distinct(cafe)
 	)


m14.1c <- stan(model_code = stan_program, data = stan_data)
```


```{r}
m14.1c
summary(m14.1c, c("Rho"))$summary
```


```{r, warning=FALSE, message=FALSE, results=FALSE}
datplot <- m14.1c %>% 
           spread_draws(Rho[i, j]) %>%
           filter(i == 1, j == 2)

ggplot(datplot, aes(Rho)) +
    geom_density() +
    xlim(-1, 1) +
    xlab('Correlation')
```

这里我用系数矩阵的方法写一遍
```{r, warning=FALSE, message=FALSE}
stan_program <- "

data {
  int n;
  int n_cafe;
  int cafe[n];
  vector[n] afternoon;
  vector[n] wait;
}
parameters {
  matrix[n_cafe, 2] a_cafe;
  real a;
  real b;
  vector<lower=0>[2] sigma_cafe;
  real<lower=0> sigma;
  corr_matrix[2] Rho;
}
transformed parameters {
  vector[2] YY[n_cafe];
  vector[2] MU;
  MU = [a, b]';
}
model {
  vector[n] mu;
  
  sigma ~ exponential(1);
  sigma_cafe ~ exponential(1);
  a ~ normal(5, 2);
  b ~ normal(-1, .5);
  Rho ~ lkj_corr(2);
  
  for (i in 1:n) {
   mu[i] = a_cafe[cafe[i], 1] + a_cafe[cafe[i], 2] * afternoon[i];
  }
  
  for (j in 1:n_cafe) {
    a_cafe[j, 1:2] ~ multi_normal(MU, quad_form_diag(Rho, sigma_cafe));
   }
  
  wait ~ normal(mu, sigma);
}
"


stan_data <- d %>% 
  tidybayes::compose_data(
 		n_cafe = n_distinct(cafe)
 	)


m14.1d <- stan(model_code = stan_program, data = stan_data)
```


```{r}
m14.1d
```


用模型矩阵+ 系数矩阵 both完成

```{r, warning=FALSE, message=FALSE}
stan_program <- "
data {
  int n;           //obs
  int K;           //num of coef including intercept
  int n_cafe;      //group deepth levels
  int cafe[n];     //group index

  vector[n] wait;  //Y variables
  matrix[n, K] X;  //model matrix
}
parameters {
  matrix[n_cafe, K] coef;
  real a;
  real b;
  vector<lower=0>[K] sigma_cafe;
  real<lower=0> sigma;
  corr_matrix[K] Rho;
}
transformed parameters {
  vector[2] MU;
  MU = [a, b]';
}
model {
  vector[n] mu;
  matrix[n, n_cafe] temp;
  temp = X * coef';
  
  for (i in 1:n) {
    mu[i] = temp[i, cafe[i]];
  }
  
  sigma ~ exponential(1);
  sigma_cafe ~ exponential(1);
  a ~ normal(5, 2);
  b ~ normal(-1, .5);
  Rho ~ lkj_corr(2);
  

  for (j in 1:n_cafe) {
    
    coef[j, 1:2] ~ multi_normal(MU, quad_form_diag(Rho, sigma_cafe));
    
  }

  wait ~ normal(mu, sigma);
}
"


stan_data <- d %>% 
  tidybayes::compose_data(
    K = 2,
    wait = wait,
    n_cafe = n_distinct(cafe),
    cafe = cafe,
    X = model.matrix(~afternoon, .)
  )


m14.1e <- stan(model_code = stan_program, data = stan_data)
```



```{r}
m14.1e
```



## 用stan-book中的方法

模型矩阵的数据 + `array of vector` 形式的系数

```{r, warning=FALSE, message=FALSE}
stan_program <- "
data {
  int N;                      // number of obs
  int K;                      // number of predictors
  matrix[N, K] X;             // model_matrix
  vector[N] y;                // y
  int J;                      // number of grouping
  int<lower=1, upper=J> g[N]; // index for grouping
}
parameters {
  vector[K] beta[J];
  vector[K] MU;
  real<lower=0> sigma;
  
  vector<lower=0>[K] tau;
  corr_matrix[K] Rho;
}

model {
  vector[N] mu;

  sigma ~ exponential(1);
  tau ~ exponential(1);
  Rho ~ lkj_corr(2);
  
  for(i in 1:N) {
    mu[i] = X[i] * beta[g[i]];  
  }
  y ~ normal(mu, sigma); 
  
  beta ~ multi_normal(MU, quad_form_diag(Rho, tau));

}
"


stan_data <- d %>% 
  tidybayes::compose_data(
    N = n,
    K = 2,
 		J = n_distinct(cafe),
    g = cafe,
    y = wait,
    X = model.matrix(~ 1 + afternoon, data = .)
 	)


m14.1f <- stan(model_code = stan_program, data = stan_data)
```


```{r}
summary(m14.1c, c("a", "b"))$summary
summary(m14.1f, c("MU"))$summary

summary(m14.1c, c("a_cafe"))$summary
summary(m14.1f, c("beta"))$summary
```




## 用brms
```{r b14.1}
b14.1 <- 
  brm(data = d, 
      family = gaussian,
      wait ~ 1 + afternoon + (1 + afternoon | cafe),
      prior = c(prior(normal(5, 2), class = Intercept),
                prior(normal(-1, 0.5), class = b),
                prior(exponential(1), class = sd),
                prior(exponential(1), class = sigma),
                prior(lkj(2), class = cor)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      seed = 867530,
      file = "fits/b14.01")
```




## Advanced varying slopes

In [Section 13.3][More than one type of cluster] we saw that data can be considered **cross-classified** if they have multiple grouping factors. We used the `chipanzees` data in that section and we only considered cross-cassification by single intercepts. Turns out cross-classified models can be extended further. Let's load and wrangle those data.


```{r}
df <- read_rds("data/chimpabzees.rds")
df
```


```{r, warning = F, message = F}
# treatment 弄成因子，事实上为了让tidybayes::compose_data()方便转化成数值，
# 最后用stan 读成 int
d <-
  df %>% 
  mutate(treatment = factor(1 + prosoc_left + 2 * condition)) #
d
```

```{r}
d %>% count(treatment)
d %>% count(actor)
d %>% count(block)
```


If I'm following along correctly with the text, McElreath's `m14.2` uses the centered parameterization. Recall from the last chapter that **brms** only supports the non-centered parameterization. Happily, McElreath's `m14.3` appears to use the non-centered parameterization. Thus, we'll skip making a `b14/2` and jump directly into making a `b14.3`. I believe one could describe the statistical model as

$$
\begin{align*}
\text{left_pull}_i & \sim \operatorname{Binomial}(n_i = 1, p_i) \\
\operatorname{logit} (p_i) & = \gamma_{\text{treatment}[i]} + \alpha_{\text{actor}[i], \text{treatment}[i]} + \beta_{\text{block}[i], \text{treatment}[i]} \\
\gamma_j & \sim \operatorname{Normal}(0, 1), \;\;\; \text{for } j = 1..4 \\
\begin{bmatrix} \alpha_{j, 1} \\ \alpha_{j, 2} \\ \alpha_{j, 3} \\ \alpha_{j, 4} \end{bmatrix} & \sim \operatorname{MVNormal} \begin{pmatrix} \begin{bmatrix} 0 \\ 0 \\ 0 \\ 0 \end{bmatrix}, \mathbf \Sigma_\text{actor} \end{pmatrix} \\
\begin{bmatrix} \beta_{j, 1} \\ \beta_{j, 2} \\ \beta_{j, 3} \\ \beta_{j, 4} \end{bmatrix} & \sim \operatorname{MVNormal} \begin{pmatrix} \begin{bmatrix} 0 \\ 0 \\ 0 \\ 0 \end{bmatrix}, \mathbf \Sigma_\text{block} \end{pmatrix} \\
\mathbf \Sigma_\text{actor} & = \mathbf{S_\alpha R_\alpha S_\alpha} \\
\mathbf \Sigma_\text{block} & = \mathbf{S_\beta R_\beta S_\beta} \\
\sigma_{\alpha, [1]}, ..., \sigma_{\alpha, [4]} & \sim \operatorname{Exponential}(1) \\
\sigma_{\beta, [1]}, ..., \sigma_{\beta, [4]}   & \sim \operatorname{Exponential}(1) \\
\mathbf R_\alpha & \sim \operatorname{LKJ}(2) \\
\mathbf R_\beta  & \sim \operatorname{LKJ}(2).
\end{align*}
$$
```{r}
glimpse(d)
```



先自己写，好像不能用block 这个在stan有特殊含义，所以改为block_id

```{r, warning=FALSE, message=FALSE}
stan_program <- "
data {
  int n;
  int n_actors;
  int n_blocks;
  int n_treatment;
  int actor[n];
  int block_id[n];
  int treatment[n];
  int pulled_left[n];
}
parameters {
    vector[n_treatment] gamma;
    vector[n_treatment] alpha[n_actors];
    vector[n_treatment] beta[n_blocks];
    vector<lower=0>[4] sigma_actor;  
    vector<lower=0>[4] sigma_block;   
    corr_matrix[4] rho_actor;
    corr_matrix[4] rho_block;
}

model {
    vector[n] p;
    gamma ~ normal(0, 1);
    sigma_actor ~ exponential(1);
    sigma_block ~ exponential(1);
    rho_actor ~ lkj_corr(4);
    rho_block ~ lkj_corr(4);
   
  for (i in 1:n) {
   p[i] = gamma[treatment[i]] + 
          alpha[actor[i], treatment[i]] + 
          beta[block_id[i], treatment[i]];
         
   p[i] = inv_logit(p[i]);
  }
  
  //alpha ~ multi_normal(rep_vector(0, 4), 
  //                    quad_form_diag(rho_actor, sigma_actor));
  //beta ~ multi_normal(rep_vector(0, 4), 
  //                    quad_form_diag(rho_block, sigma_block));

  for(j in 1:n_actors) {
    alpha[j] ~ multi_normal(rep_vector(0, 4), quad_form_diag(rho_actor, sigma_actor));
  }
    for(j in 1:n_blocks){
      beta[j] ~ multi_normal(rep_vector(0, 4), quad_form_diag(rho_block, sigma_block));
    }
  
  
  pulled_left ~ binomial(1, p); 
}
"

stan_data <- d %>%
  rename(block_id = block) %>%
  tidybayes::compose_data(
    	n_actors = n_distinct(actor),
    	n_blocks = n_distinct(block_id),
    	n_treatment = n_distinct(treatment)
  )


m14.2 <- stan(model_code = stan_program, data = stan_data)
```



```{r}
m14.2
```



用矩阵matrix装系数也是一样的效果

```{r, warning=FALSE, message=FALSE}
stan_program <- "
data {
  int n;
  int n_actors;
  int n_blocks;
  int n_treatment;
  int actor[n];
  int block_id[n];
  int treatment[n];
  int pulled_left[n];
}
parameters {
    vector[n_treatment] gamma;
    
    matrix[n_actors, n_treatment] alpha;
    matrix[n_blocks, n_treatment] beta;
    

    vector<lower=0>[n_treatment] sigma_actor;  
    vector<lower=0>[n_treatment] sigma_block;   

    corr_matrix[n_treatment] rho_actor;
    corr_matrix[n_treatment] rho_block;
}

model {
    vector[n] p;
    
    gamma ~ normal(0,1);
    sigma_actor ~ exponential(1); 
    sigma_block ~ exponential(1);
    rho_actor ~ lkj_corr(2);
    rho_block ~ lkj_corr(2);
    
   
   for (i in 1:n) {
   p[i] = gamma[treatment[i]] + 
          alpha[actor[i], treatment[i]] + 
          beta[block_id[i], treatment[i]];
         
   p[i] = inv_logit(p[i]);
  }
  
   for(j in 1:n_actors) {
    alpha[j] ~ multi_normal(rep_vector(0, 4), quad_form_diag(rho_actor, sigma_actor));
  }
    for(j in 1:n_blocks){
      beta[j] ~ multi_normal(rep_vector(0, 4), quad_form_diag(rho_block, sigma_block));
  }
     
  pulled_left ~ binomial(1, p); 
}
"

stan_data <- d %>%
  rename(block_id = block) %>%
  tidybayes::compose_data(
    	n_actors = n_distinct(actor),
    	n_blocks = n_distinct(block_id),
    	n_treatment = n_distinct(treatment)
  )


m14.2a <- stan(model_code = stan_program, data = stan_data)
```

```{r}
m14.2a
```





## 模型14.3
这个的数学公式在哪里？
```{r, warning=FALSE, message=FALSE, results=FALSE}
stan_program <- "
data {
  int n;
  int n_actors;
  int n_blocks;
  int n_treatment;
  int actor[n];
  int block_id[n];
  int treatment[n];
  int pulled_left[n];
}
parameters {
   matrix[n_treatment, n_actors] z_actor;
   matrix[n_treatment, n_blocks] z_block;
   vector[n_treatment] g; 
   vector<lower=0>[n_treatment] sigma_actor;
   vector<lower=0>[n_treatment] sigma_block;
   cholesky_factor_corr[n_treatment] L_Rho_block;
   cholesky_factor_corr[n_treatment] L_Rho_actor;
}
transformed parameters {
   matrix[n_actors, n_treatment] alpha;
   matrix[n_blocks, n_treatment] beta;
   beta = (diag_pre_multiply(sigma_block, L_Rho_block) * z_block)';
   alpha = (diag_pre_multiply(sigma_actor, L_Rho_actor) * z_actor)';
}
model{
    vector[n] p;
    L_Rho_block ~ lkj_corr_cholesky( 2 );
    sigma_block ~ exponential( 1 );
    L_Rho_actor ~ lkj_corr_cholesky( 2 );
    sigma_actor ~ exponential( 1 );
    g ~ normal( 0 , 1 );
    to_vector( z_block ) ~ normal( 0 , 1 );
    to_vector( z_actor ) ~ normal( 0 , 1 );
    for ( i in 1:n ) {
        p[i] = g[treatment[i]] + alpha[actor[i], treatment[i]] + beta[block_id[i], treatment[i]];
        p[i] = inv_logit(p[i]);
    }
    pulled_left ~ binomial( 1 , p );
}
"


stan_data <- d %>%
  rename(block_id = block) %>%
  tidybayes::compose_data(
    	n_actors = n_distinct(actor),
    	n_blocks = n_distinct(block_id),
    	n_treatment = n_distinct(treatment)
  )



m14.3 <- stan(model_code = stan_program, data = stan_data)
```


```{r}
m14.3
```




brms版本
```{r, warning = F, message = F}
# wrangle.
d1 <-
  df %>% 
  mutate(actor     = factor(actor),
         block     = factor(block),
         treatment = factor(1 + prosoc_left + 2 * condition),
         # this will come in handy, later
         labels    = factor(treatment,
                            levels = 1:4,
                            labels = c("r/n", "l/n", "r/p", "l/p")))
d1
```


```{r b14.3}
b14.3 <- 
  brm(data = d1, 
      family = binomial,
      pulled_left | trials(1) ~ 0 + treatment + (0 + treatment | actor) + (0 + treatment | block),
      prior = c(prior(normal(0, 1), class = b),
                prior(exponential(1), class = sd, group = actor),
                prior(exponential(1), class = sd, group = block),
                prior(lkj(2), class = cor, group = actor),
                prior(lkj(2), class = cor, group = block)),
      iter = 2000, warmup = 1000, chains = 4, cores = 4,  
      seed = 4387510,
      file = "fits/b14.03")
```


```{r}
b14.3 
```

